---
title: "Statistical Rethinking - Exercises from  Chapter 2"
author: "Hugo Mailhot"
date: "October 9, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Medium


### 2M1 
*Recall the globe tossing model from the chapter. Compute and plot the grid approximate posterior distribution for each of the following sets of observations. In each case, assume a uniform prior for _p_.*

```{r grid_computing}
grid_approx <- function(p_grid, prior, k, n){
    # compute likelihood at each value in grid
    likelihood <- dbinom( k , size=n , prob=p_grid )
    # compute product of likelihood and prior
    unstd.posterior <- likelihood * prior
    # standardize the posterior, so it sums to 1
    posterior <- unstd.posterior / sum(unstd.posterior)
    return(posterior)
}

# Since we are going to use the same code to plot each time,
# wrap everything up in a single function.
approx_and_plot <- function(p_grid, prior, k, n, points){
    posterior <- grid_approx(p_grid, prior, k, n)
    plot( p_grid , posterior , type="b" ,
          xlab="probability of water" , ylab="posterior probability" )
    mtext( paste(points, " points" ))
}
```

Here we define our grid and uniform prior, which we'll use in the next 3 approximations.

```{r uniform_prior}
points <- 30
p_grid <- seq( from=0 , to=1 , length.out=points )
uni_prior <- rep(1, points)
```

#### 1) W, W, W


```{r 2M1.1}
approx_and_plot(p_grid, uni_prior, k=3, n=3, points)
```

#### 2) W, W, W, L

```{r 2M1.2}
approx_and_plot(p_grid, uni_prior, k=3, n=4, points)
```

#### 3) L, W, W, L, W, W, W

```{r 2M1.3}
approx_and_plot(p_grid, uni_prior, k=5, n=7, points)
```

### 2M2 
*Now assume a prior for p that is equal to zero when p < 0.5 and is a positive constant when p ≥ 0.5. Again compute and plot the grid approximate posterior distribution for each of the sets of observations in the problem just above.*

Define the new prior

```{r step_prior}
step_prior <-rep( c(0,1), each = points/2)
```

#### 1) W, W, W


```{r 2M2.1}
approx_and_plot(p_grid, step_prior, k=3, n=3, points)
```

#### 2) W, W, W, L

```{r 2M2.2}
approx_and_plot(p_grid, step_prior, k=3, n=4, points)
```

#### 3) L, W, W, L, W, W, W

```{r 2M2.3}
approx_and_plot(p_grid, step_prior, k=5, n=7, points)
```

## 2M3

*Suppose there are two globes, one for Earth and one for Mars. The Earth globe is 70% covered in water. The Mars globe is 100% land. Further suppose that one of these globes—you don’t know which was tossed in the air and produced a “land” observation. Assume that each globe was equally likely to be tossed. Show that the posterior probability that the globe was the Earth, conditional on seeing “land” (Pr(Earth|land)), is 0.23.*

First use Bayes' Theorem:
$$P(Earth|land) = \frac{P(land|Earth)P(Earth)}{P(land)}$$

Now show the following:
$$\frac{P(land|Earth)P(Earth)}{P(land)} = 0.23$$

From the problem description we know the following:
$$P(Earth) = 0.5$$
$$P(land|Earth) = 1 - 0.7 = 0.3$$
$$P(land) = (1 + 0.3)/2 = 0.65$$

This last one is because without knowing which planet we have, the probability of land is the average probability of land between the two of them.

Putting everything together:
$$P(Earth|land) = \frac{0.3 \times 0.5}{0.65} \approx 0.23 $$